One-arm spirals in locally isothermal, radially structured self-gravitating discs

### 2014-11-02 paper/discussion.tex
added variable q discussion

We now invoke local theory by setting $d/dR \to \ii k$ where $k$ is
real, and using the local solution to the Poisson equation
\begin{align}
  \delta\Phi_m = -\frac{2\pi G}{|k|}\delta\Sigma_m 
\end{align}
\citep{shu91}. Then the radial velocity perturbation becomes
\begin{align}
  \delta v_{Rm} =
  \frac{1}{D}\frac{\delta\Sigma_m}{\Sigma}\left(\frac{2\pi G
      \Sigma}{|k|} - c_s^2\right)\left(k\sbar + \frac{2\ii
      m\Omega}{R}\right),
\end{align}
so that 
\begin{align}
  \ii\delta\Sigma_m \left(\frac{\delta v_{Rm}}{\sbar}\right)^* =
  \frac{1}{D^*}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
      \Sigma}{|k|} - c_s^2\right)\left(\ii k  + \frac{2
      m\Omega}{R\sbar^*}\right). 
\end{align}
For a neutral mode, $\sbar$, and therfore $D$ is real. Then, after
taking the imaginary part of the above expression, the torque
density is 
\begin{align}
  T_\mathrm{BG} &=
  \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{D}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
      \Sigma}{|k|} - c_s^2\right)\right] \notag\\
&\simeq \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{\left(\kappa^2 -
        m^2\Omega^2\right)}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi
      G \Sigma}{|k|} - c_s^2\right)\right]. 
\end{align}
where the second line follows from the low-frequency approximation
$|\sigma|\ll m\Omega$.

To determine the sign of $T_\mathrm{BG}$ for low frequency one-arm spirals,
we examine each factor separately below for $m=1$:
\begin{itemize}
\item We observe trailing spirals in our simulations, so $k>0$. 
\item The factor $\kappa^2 - \Omega^2>0$ for our self-gravitating disc
  models in the region where the spiral appears. 
\item We measure $|k|\sim \pi G\Sigma/c_s^2$, so the last factor in
  the square brackets is positive. Note that this factor is positive
  provided $|k| < 2/HQ$ or wavelengths $\lambda > \pi Q H$.  
\end{itemize}
These considerations imply the factor in the square brackets is
positive. Then, since $dc_s^2/dR < 0$, $T_\mathrm{BG}<0$ so the background
disc applies a negative torque on the low-frequency $m=1$ spiral,
which has negative angular momentum. This suggests the spiral
amplitude will grow.

>>>>

We now invoke local theory by setting $d/dR \to \ii k$ where $k$ is
real, and using the local solution to the Poisson equation
\begin{align}
  \delta\Phi_m = -\frac{2\pi G}{|k|}\delta\Sigma_m 
\end{align}
\citep{shu91}. Then the radial velocity perturbation becomes
\begin{align}
  \delta v_{Rm} =
  \frac{1}{D}\frac{\delta\Sigma_m}{\Sigma}\left(\frac{2\pi G
      \Sigma}{|k|} - c_s^2\right)\left(k\sbar + \frac{2\ii
      m\Omega}{R}\right),
\end{align}
so that 
\begin{align}
  \ii\delta\Sigma_m \left(\frac{\delta v_{Rm}}{\sbar}\right)^* =
  \frac{1}{D^*}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
      \Sigma}{|k|} - c_s^2\right)\left(\ii k  + \frac{2
      m\Omega}{R\sbar^*}\right). 
\end{align}
For a neutral mode, $\sbar$, and therfore $D$ is real. Then, after
taking the imaginary part of the above expression, the torque
density is 
\begin{align}\label{tbg_explicit}
  T_\mathrm{BG} &=
  \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{D}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
      \Sigma}{|k|} - c_s^2\right)\right].
% &\simeq \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{\left(\kappa^2 -
%         m^2\Omega^2\right)}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi
%       G \Sigma}{|k|} - c_s^2\right)\right]. 
\end{align}
%where the second line follows from the low-frequency approximation
%$|\sigma|\ll m\Omega$. 
% To determine the sign of $T_\mathrm{BG}$ for low frequency one-arm spirals,
% we examine each factor separately below for $m=1$:
% \begin{itemize}
% \item We observe trailing spirals in our simulations, so $k>0$. 
% \item The factor $\kappa^2 - \Omega^2>0$ for our self-gravitating disc
%   models in the region where the spiral appears. 
% \item We measure $|k|\sim \pi G\Sigma/c_s^2$, so the last factor in
%   the square brackets is positive. Note that this factor is positive
%   provided $|k| < 2/HQ$ or wavelengths $\lambda > \pi Q H$.  
% \end{itemize}
We now use the local dispersion relation,
Eq. \ref{dispersion}, to make the replacement $D = 2\pi G \Sigma |k| -
k^2c_s^2$ in the above expression. This gives
\begin{align}\label{tbg_simple}
  T_\mathrm{BG} =
  \frac{m}{2}\frac{dc_s^2}{dR}\frac{1}{k}\frac{|\delta\Sigma_m|^2}{\Sigma}.  
\end{align} 
This torque density is negative for trailing waves in discs with
temperature decreasing outwards ($k>0$ and $dc_s^2/dR<0$,
respectively). Note that this conclusion does not rely on the
low-frequency approximation and holds for any $m$.

However, if the linear disturbance under consideration \emph{is} a
low-frequency $m=1$ 
trailing spiral wave, then it has negative angular
momentum. If $dc_s^2/dR<0$, as is typical for circumstellar or
protoplanetary discs, then $T_\mathrm{BG}<0$, and the background disc
applies a negative torque on the disturbance, which further decreases
its angular momentum. This suggests the spiral amplitude will grow.

%torque effective for large scales or sharp gradients

%Then, since $dc_s^2/dR < 0$, $T_\mathrm{BG}<0$ so the background
%disc applies a negative torque on the low-frequency $m=1$ spiral,
%which has negative angular momentum. This suggests the spiral
%amplitude will grow.

\subsection{Role of disc structure}

### 2014-11-02 paper/model.tex
added analytical discussion
added variable q discussion

\section{Governing equations and numerical models}\label{model} 
The system of interest is a self-gravitating, inviscid fluid disc
orbting a central star of mass $M_*$. We will mainly consider two-dimensional
(2D, or razor-thin) discs in favour of numerical
resolution, but have also carried out some three-dimensional (3D) disc
simulations. Hence, for generality we describe the system in
3D. We use both cylindrical $(R,\phi,z)$ and spherical polar
coordinates $(r,\theta,\phi)$ centred on the star. 
The governing fluid equations in 3D are  
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) =
  0,\label{cont_eq}\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bv = -\frac{1}{\rho}\nabla
  p -\nabla \Phi_\mathrm{tot}\label{mom_eq},\\ 
  & \nabla^2\Phi_d = 4 \pi G \rho \label{poisson}, 
\end{align}
where $\rho$ is the mass density, $\bm{v}=(v_r,v_\theta,v_\phi)$ the
fluid velocity, $p$ is the pressure and the total potential
$\Phi_\mathrm{tot} = \Phi_* + \Phi_d$ consists of that from the
central star, 
\begin{align}
  \Phi_*(r) = -\frac{GM_*}{r}, 
\end{align}
where $G$ is the gravitational constant,  and the disc potential
$\Phi_d$. We assume a locally 
isothermal equation of state 
\begin{align}
  p = c_s^2(R)\rho,
\end{align}
where the sound-speed is given by 
\begin{align}
  c_s(R) = hR_0\Omega_k(R_0)\left(\frac{R}{R_0}\right)^{-q/2}, 
\end{align}
where $h$ is the disc aspect-ratio $R=R_0$, 
$\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency, and
$q$ is the imposed temperature gradient. A strictly isothermal disc
corresponds to $q=0$, and a disc with constant aspect-ratio
corresponds to $q=1$.

@@

\subsection{Disc model and initial conditions}
%We consider a smooth disc which contains a surface density bump
%between $R\in[R_{d1},R_{d2}]$, mimicking mass built-up in a dead
%zone. 
We adopt a modified power-law disc with surface
density profile 
\begin{align}
  \Sigma(R) = \Sigma_\mathrm{ref} \left(\frac{R}{R_0}\right)^{-s}\times B(R;
  R_{d1}, R_{d2}, \epsilon, \delta R_d), 
\end{align}
where $R_0$ is a reference cylindrical radius and $s$ is the power-law
index describing the smooth disc, and $\Sigma_\mathrm{ref}$ is a
surface density scale chosen by specifying $Q_\mathrm{out}$,
the Keplerian Toomre parameter at $R=R_{d2}$,
\begin{align}
  Q_\mathrm{out} = \left.\frac{c_s\Omega_k}{\pi G
    \Sigma}\right|_{R=R_{d2}}. 
\end{align}

The bump function
$B(R)$ represents a surface density boost between
$R\in[R_{d1},R_{d2}]$ by a factor $\epsilon^{-1}>1$,
and $\delta R_d$ is the transition width between the bump and the
smooth disc. We choose 
\begin{align}
  &B(R) = f_1(R)\times f_2(R),\\
  &f_1(R) = \frac{1}{2}\left(1 - \epsilon\right)\left[1 +
    \tanh\left(\frac{R-R_{d1}}{\Delta_1}\right)\right]  + \epsilon,\\
  &f_2(R) = \frac{1}{2}\left(1 - \epsilon\right)\left[1 -
    \tanh\left(\frac{R-R_{d2}}{\Delta_2}\right)\right]  + \epsilon,
\end{align}
where $\Delta_1 = \delta R_d \times H(R_{d1})$ and similarly for
$R_{d2}$. The bump function mimicks mass built-up in a dead zone.

@@

Our fiducial parameter values are: $s=2$, $R_{d1}=R_0$, $R_{d2}=2R_0$,
$\epsilon=0.1$, $\delta R_d=5$, $h=0.05$ and
$Q_\mathrm{out}=2$. An example of the initial surface density and the
Toomre parameter (see\S\ref{wkb}) is shown in Fig. \ref{initial_surf}. 
Initially there is no vertical or radial velocity
($v_R = v_r = v_\theta = 0$). The azimuthal velocity is initialized to
satisfy centrifugal balance with pressure and gravity,
\begin{align}
  \frac{v_\phi^2}{r} = \frac{1}{\rho}\frac{\p p}{\p r} + \frac{\p
    \Phi_\mathrm{tot}}{\p r}
\end{align}
and similarly in 2D. The angular velocity is $\Omega = v_\phi/R$.

>>>>

\section{Governing equations and numerical models}\label{model} 
The system of interest is a self-gravitating, inviscid fluid disc
orbting a central star of mass $M_*$. We will mainly consider two-dimensional
(2D, or razor-thin) discs in favour of numerical
resolution, but have also carried out some three-dimensional (3D) disc
simulations. Hence, for generality we describe the system in
3D. We use both cylindrical $(R,\phi,z)$ and spherical polar
coordinates $(r,\theta,\phi)$ centred on the star. 
The governing fluid equations in 3D are  
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) =
  0,\label{cont_eq}\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bv = -\frac{1}{\rho}\nabla
  p -\nabla \Phi_\mathrm{tot}\label{mom_eq},\\ 
  & \nabla^2\Phi_d = 4 \pi G \rho \label{poisson}, 
\end{align}
where $\rho$ is the mass density, $\bm{v}=(v_r,v_\theta,v_\phi)$ the
fluid velocity, $p$ is the pressure and the total potential
$\Phi_\mathrm{tot} = \Phi_* + \Phi_d$ consists of that from the
central star, 
\begin{align}
  \Phi_*(r) = -\frac{GM_*}{r}, 
\end{align}
where $G$ is the gravitational constant,  and the disc potential
$\Phi_d$. We impose a locally 
isothermal equation of state 
\begin{align}
  p = c_s^2(R)\rho,
\end{align}
where the sound-speed is given by 
\begin{align}\label{sound-speed}
  c_s(R) = hR_0\Omega_k(R_0)\left(\frac{R}{R_0}\right)^{-q/2}, 
\end{align}
where $h$ is the disc aspect-ratio $R=R_0$, 
$\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency, and
$q$ is the imposed temperature gradient. A strictly isothermal disc
corresponds to $q=0$, and a disc with constant aspect-ratio
corresponds to $q=1$.

@@

\subsection{Disc model and initial conditions}
%We consider a smooth disc which contains a surface density bump
%between $R\in[R_{1},R_{2}]$, mimicking mass built-up in a dead
%zone. 
We adopt a modified power-law disc with surface
density profile 
\begin{align}
  \Sigma(R) = \Sigma_\mathrm{ref} \left(\frac{R}{R_0}\right)^{-s}\times B(R;
  R_{1}, R_{2}, \epsilon, \delta R_d), 
\end{align}
where $R_0$ is a reference cylindrical radius and $s$ is the power-law
index describing the smooth disc, and $\Sigma_\mathrm{ref}$ is a
surface density scale chosen by specifying $Q_\mathrm{out}$,
the Keplerian Toomre parameter at $R=R_{2}$,
\begin{align}
  Q_\mathrm{out} = \left.\frac{c_s\Omega_k}{\pi G
    \Sigma}\right|_{R=R_{2}}. 
\end{align}

The bump function
$B(R)$ represents a surface density boost between
$R\in[R_{1},R_{2}]$ by a factor $\epsilon^{-1}>1$,
and $\delta R_d$ is the transition width between the bump and the
smooth disc. We choose 
\begin{align}
  &B(R) = f_1(R)\times f_2(R),\\
  &f_1(R) = \frac{1}{2}\left(1 - \epsilon\right)\left[1 +
    \tanh\left(\frac{R-R_{1}}{\Delta_1}\right)\right]  + \epsilon,\\
  &f_2(R) = \frac{1}{2}\left(1 - \epsilon\right)\left[1 -
    \tanh\left(\frac{R-R_{2}}{\Delta_2}\right)\right]  + \epsilon,
\end{align}
where $\Delta_1 = \delta R_d \times H(R_{1})$ and similarly for
$R_{2}$. The bump function mimicks mass built-up in a dead zone.

@@

Our fiducial parameter values are: $s=2$, $R_{1}=R_0$, $R_{2}=2R_0$,
$\epsilon=0.1$, $\delta R_d=5$, $h=0.05$ and
$Q_\mathrm{out}=2$. An example of the initial surface density and the
Toomre parameter (see\S\ref{wkb}) is shown in Fig. \ref{initial_surf}. 
Initially there is no vertical or radial velocity
($v_R = v_r = v_\theta = 0$). The azimuthal velocity is initialized to
satisfy centrifugal balance with pressure and gravity,
\begin{align}
  \frac{v_\phi^2}{r} = \frac{1}{\rho}\frac{\p p}{\p r} + \frac{\p
    \Phi_\mathrm{tot}}{\p r}
\end{align}
and similarly in 2D. The angular velocity is $\Omega = v_\phi/R$.

### 2014-11-02 paper/results.tex
added variable q discussion

\section{Results for two-dimensional discs}
We first present results from FARGO simulations. The 2D disc spans
$[R_\mathrm{min}, R_\mathrm{max}] = [0.4,10]R_0$. This gives a total
disc mass $M_{d}=0.086M_*$. The mass within
$R\in[R_\mathrm{min},R_{d1}]$ is $0.017M_*$, that within
$R\in[R_{d1},R_{d2}]$ is $0.049M_*$, and that within
$R\in[R_{d2},R_\mathrm{max}]$ is $0.021M_*$. We use a resolution of
$N_R\times N_\phi = 1024\times 2048$, or about $16$ grids per $H$, and
adopt $\epsilon_g=10^{-4}H$ for the   
self-gravity softening length\footnote{In 2D self-gravity, $\epsilon_g$ also
  approximates for the vertical disc thickness, so a more appropriate
  value would be $\epsilon_g\sim H$ \citep{muller12}. However, because
  $\epsilon_g\propto R$ is needed in FARGO, the Poisson kernel
  (Eq. \ref{2d_grav}) is no longer symmetric in $(R,R^\prime)$. We
  choose a small  
  $\epsilon_g$ in favour of angular momentum conservation, keeping in
  mind that the strength of self-gravity will be over-estimated.}.

\subsection{Reference run with random perturbations}
In this fiducial simulation we subject the disc to initial perturbations in
cylindrical radial velocity, 
\begin{align}\label{randpert}
  v_R \to v_R+ c_s\frac{\delta}{M}
  \exp{\left[-\frac{1}{2}\left(\frac{R-\overline{R}_d}{\Delta 
          R_d}\right)^2\right]}\sum_{m=1}^M\cos{m\phi},
\end{align}
where $\delta$ is an amplitude, $\overline{R}_d = (R_{d1}+R_{d2})/2$
and $\Delta R_d = (R_{d2}-R_{d1})/2$. We use $M=10$ and choose 
$\delta\in[-10^{-3},10^{-3}]$ randomly but independent of $\phi$.

Fig. \ref{fargo_modeamp} plots evolution of the maximum
non-axisymmetric surface density amplitudes in $R\in[R_{d1},R_{d2}]$
for $m\in[1,10]$. Snapshots from the simulation are shown in
Fig. \ref{fargo_2d}. 
At early times $t\lesssim100P_0$ the disc 
dominated by low-amplitude high-$m$ perturbations. The $m\geq4$ modes
growth initially and saturate (or decays) after $t=40P_0$. Notice the
low $m\leq 2$ modes decay initially, but grows between $t\in[20,40]P_0$,
possibily due to saturation of the high-$m$ modes  
\citep{laughlin96,laughlin97}. However, the $m=1$ mode begins to grow
again after $t=70P_0$, and eventually dominates the annulus.

@@

By examining the disc evolution in $t\in[80,120]P_0$, we find the
one-arm spiral in Fig. \ref{2d_fargo_viz} has a 
co-rotation radius and growth rate of
\begin{align*}
  &R_c \simeq 4.4R_0,\\
  &\gamma\simeq 0.014\Omega_k(R_0) = 0.13\Omega_k(R_c). 
\end{align*}
The spiral can be considered as low frequency because $|\omega|
\lesssim 0.3\Omega$ (recall $\omega = 
m\Omega_p$) in the region it is promininent $R\in[R_{d1},R_{d2}]$. The
growth rate is also slow relative to the local rotation. This gives a
nearly stationary spiral pattern.

@@

In order to track down the origin for the (slow) growth of the
$m=1$ spiral, we invoke angular momentum conservation for linear
perturbations, Eq. \ref{lin_ang_mom_cons}. Assuming fluxes are
negligible at the disc boundaries, we can integrate
Eq. \ref{lin_ang_mom_cons} to give
\begin{align}\label{baroclinic_torque_int}
  \frac{d}{dt}\underbrace{\int_{R_\mathrm{min}}^{R_\mathrm{max}}\jlin
    2\pi R dR}_{\jlintot} 
  =\int_{R\mathrm{min}}^{R\mathrm{max}}T_\mathrm{BG} 2\pi R dR, 
\end{align}
where we recall $T_{BG}$ is the torque density related to the imposed
sound-speed profile (Eq. \ref{baroclinic_torque}). We have explicitly
computed both sides of Eq. \ref{baroclinic_torque_int} using
simulation data, and compare them in Fig. \ref{fargo_angmom_ex}.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_ang_fargo.ps} 
  \caption{
    \label{fargo_angmom_ex}} 
\end{figure}

@@

>>>>

\section{Results}
We first present results from FARGO simulations. The 2D disc spans
$[R_\mathrm{min}, R_\mathrm{max}] = [0.4,10]R_0$. This gives a total
disc mass $M_{d}=0.086M_*$. The mass within
$R\in[R_\mathrm{min},R_{1}]$ is $0.017M_*$, that within
$R\in[R_{1},R_{2}]$ is $0.049M_*$, and that within
$R\in[R_{2},R_\mathrm{max}]$ is $0.021M_*$. We use a resolution of
$N_R\times N_\phi = 1024\times 2048$, or about $16$ grids per $H$, and
adopt $\epsilon_g=10^{-4}H$ for the   
self-gravity softening length\footnote{In 2D self-gravity, $\epsilon_g$ also
  approximates for the vertical disc thickness, so a more appropriate
  value would be $\epsilon_g\sim H$ \citep{muller12}. However, because
  $\epsilon_g\propto R$ is needed in FARGO, the Poisson kernel
  (Eq. \ref{2d_grav}) is no longer symmetric in $(R,R^\prime)$. We
  choose a small  
  $\epsilon_g$ in favour of angular momentum conservation, keeping in
  mind that the strength of self-gravity will be over-estimated.}.

\subsection{Reference run with random perturbations}
In this fiducial simulation we subject the disc to initial perturbations in
cylindrical radial velocity, 
\begin{align}\label{randpert}
  v_R \to v_R+ c_s\frac{\delta}{M}
  \exp{\left[-\frac{1}{2}\left(\frac{R-\overline{R}_d}{\Delta 
          R_d}\right)^2\right]}\sum_{m=1}^M\cos{m\phi},
\end{align}
where $\delta$ is an amplitude, $\overline{R}_d = (R_{1}+R_{2})/2$
and $\Delta R_d = (R_{2}-R_{1})/2$. We use $M=10$ and choose 
$\delta\in[-10^{-3},10^{-3}]$ randomly but independent of $\phi$.

Fig. \ref{fargo_modeamp} plots evolution of the maximum
non-axisymmetric surface density amplitudes in $R\in[R_{1},R_{2}]$
for $m\in[1,10]$. Snapshots from the simulation are shown in
Fig. \ref{fargo_2d}. 
At early times $t\lesssim100P_0$ the disc 
dominated by low-amplitude high-$m$ perturbations. The $m\geq4$ modes
growth initially and saturate (or decays) after $t=40P_0$. Notice the
low $m\leq 2$ modes decay initially, but grows between $t\in[20,40]P_0$,
possibily due to saturation of the high-$m$ modes  
\citep{laughlin96,laughlin97}. However, the $m=1$ mode begins to grow
again after $t=70P_0$, and eventually dominates the annulus.

@@

By examining the disc evolution in $t\in[80,120]P_0$, we find the
one-arm spiral in Fig. \ref{2d_fargo_viz} has a 
co-rotation radius and growth rate of
\begin{align*}
  &R_c \simeq 4.4R_0,\\
  &\gamma\simeq 0.014\Omega_k(R_0) = 0.13\Omega_k(R_c). 
\end{align*}
The spiral can be considered as low frequency because $|\omega|
\lesssim 0.3\Omega$ (recall $\omega = 
m\Omega_p$) in the region it is promininent $R\in[R_{1},R_{2}]$. The
growth rate is also slow relative to the local rotation. This gives a
nearly stationary spiral pattern.

@@

In order to track down the origin for the (slow) growth of the
$m=1$ spiral, we invoke angular momentum conservation for linear
perturbations, Eq. \ref{lin_ang_mom_cons}. Assuming fluxes are
negligible at the disc boundaries, we can integrate
Eq. \ref{lin_ang_mom_cons} to give
\begin{align}\label{baroclinic_torque_int}
  \frac{d}{dt}\underbrace{\int_{R_\mathrm{min}}^{R_\mathrm{max}}\jlin
    2\pi R dR}_{\jlintot} 
  =\int_{R\mathrm{min}}^{R\mathrm{max}}T_\mathrm{BG} 2\pi R dR, 
\end{align}
where we recall $T_{BG}$ is the torque density related to the imposed
sound-speed profile (Eq. \ref{baroclinic_torque}). We have explicitly
computed both sides of Eq. \ref{baroclinic_torque_int} using
simulation data, and compare them in Fig. \ref{fargo_angmom_ex}. There
is a good match between the two torques, especially at early times
$t\lesssim110P_0$. The match is less good later on, when the spiral
amplitude is no longer small ($\Delta\Sigma_1\sim 0.2$ at $t=110P_0$
and $\Delta\Sigma_1\sim 0.4$ by $t=120P_0$) and linear theory becomes
less valid. Fig. \ref{fargo_angmom_ex} confirms that angular momentum
is lost from the wave to the disc.

%wave losses at boundaries

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_ang_fargo.ps} 
  \caption{Rate of change of the $m=1$ wave angular momentum as defined by
    Eq. \ref{baroclinic_torque_int} (solid) compared to the torque
    exerted on the wave associated with the background temperature
    gradient (dotted). 
    \label{fargo_angmom_ex}} 
\end{figure}

@@

\subsection{Dependence on the imposed temperature gradient} 
At this point we hypothesize the growth of the $m=1$ spiral mode is
associated with the imposed temperature gradient. The torque
responsible for its growth, Eq. \ref{baroclinic_torque}, is
proportional to $q$, the power-law index used to set the sound-speed
profile (Eq. \ref{sound-speed}). To test this, we carried out
simulations with $q\in[0,1]$ using only an $m=1$ initial 
perturbation. However, to keep the Toomre profiles similar, we adjust
the surface density power-law index such that $s = (3+q)/2$ for each
case presented below. %should not be important, local effects

Fig. \ref{fargo_varq} compares the $m=1$ spiral amplitudes as a
function of $q$. We indeed observe slower growth with decreasing
$q$. Although the figure indicates growth for the strictly isothermal
disc ($q=0$), we did not observe a coherent one-arm spiral upon
inspection of the $m=1$ surface density field. The growth in this case
may be associated with higher $m$ modes. %"non-linear effects"

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_fargo_varq.ps}   
  \caption{Evolution of the $m=1$ spiral amplitude as a function of
    the imposed sound-speed gradient $q$. The maximum value of the
    $m=1$ surface density in $R\in[R_1,R_2]$ is shown. 
    \label{fargo_varq}} 
\end{figure}

We plot growth rates of the $m=1$ spiral as a function of $q$ in 
Fig. \ref{fargo_varq_growth}. The correlation is well-fitted by a
linear relation
\begin{align*}
  \gamma = 0.014 q \Omega_k{R_0}. 
\end{align*}
This confirms that the temperature gradient is responsible for the
development of the one-arm spiral observed in our simulations. 
\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ratemax_fargo_varq.ps}    
  \caption{Growth rates of the $m=1$ spiral mode as a function of the
    imposed sound-speed gradient $q$ (asterisks). A linear fit is also
    plotted (dotted line). 
    \label{fargo_varq_growth}} 
\end{figure}

% only cases where a coherent spiral is seen is plotted
